Efficient Planning for Sokoban with Answer Set Programming - A Case Study

### 2018-09-18 BATexFiles/Draft.tex
Textversion 1 abgeschlossen

%\titleformat{\chapter}
%  {\normalfont\LARGE\bfseries}{\thechapter}{1em}{}

@@

\section{Rules}

@@

IM DOING ASP AND THAT IS DIFFERENT TO MOST OTHER APPROACHES I LOOKED AT

@@

For illustration consider the level shown in Figure \ref{fig:levelVis}. The corresponding fact representation can be seen in Listing \ref{lst:lvl1}. The board size is fixed by the constants $dimX$ and $dimY$; the origin (1,1) is in the top left corner. The positions of player and boxes are stored in the at/3 predicate. The first argument represents the identifier. A zero stands for the player, while all boxes have numbers greater than zero. The second and third argument are the X and Y coordinates, signifying the initial position of the object. For instance init(at(0,1,2)) means the player is at the initial position (1,2). Because I am not exclusively looking at rectangular levels I decided to store all legal fields (meaning fields where a box or player can be placed) by using the predicate field/2. In rectangular levels the legal fields can be defined in one line. Here init(field(1..5,1..4)) (l.4) means the crossproduct of all the numbers in the pools (i.e. (1,1),..,(5,1),(1,2),..,(5,4)).

@@

\subsection{Split Encoding}
\lstinputlisting{../EncodingsClean/SokobanSplitEncoding.lp}
\begin{lstlisting}[caption={Split Encoding},label=list:SplitEnc]
\end{lstlisting}
The split encoding uses a technique common in ASP planning. When all coordinates are split into a vertical and a horizontal information the grounder does not have to ground all combinations of coordinates ($X * Y$) but only the sum of coordinates ($X + Y$). This has the advantage of less ground instances, but the disadvantage of a harder solving process. Just compare the ramifications in the plain encoding those in the split encoding. The plain encoding can deal with all ramifications in just three rules, while the split encoding needs ten rules. Obviously when splitting the coordinates into horizontal and vertical the rules deriving the new coordinates have to be twice as many. The other problem is that it is harder to find out whether two objects meet each other at a time. That is why the frame axiom, signifying that boxes that are not touched do not move, has six lines, where it had only one line before. This is why this encoding has a performance problem.
The last few changes are a second line for the initial readout of the level in lines 4 and 5, a few updates to the integrity constraints in lines 24 and 25 and a slightly different goal definition using the new coordinates.
%Here is another encoding, following a common ASP planning technique that aims to reduce ground instances. By splitting the horizontal and vertical coordinates, the program does not have to ground all combinations ($dimX * dimY$) of coordinates, but just each coordinate by itself (totaling $dimX + dimY$).% only grounds $(dimX + dimY) * horizon * N$ different positions instead of $dimX * dimY * horizon * N$.

@@

Whenever an AI tries to solve a planning problem with a naive encoding, it always has to guess all possible combinations of moves and just has to play them through. There are a lot of moves that instantly destroy any chanco in ever satisfiying the level afterwards. Those gamestates from which no solution can be found are called deadlocks. A search tree can be greatly reduced if one were to remove all deadlocked branches. The difficulty is that some deadlocks are very hard to notice and sometimes it takes more time elimintaing the deadlock than to explore the search tree without eliminating. This is why the deadlock detection approach takes only obvious deadlocks. There are three different approaches I tried to remove deadlocks. The Forbidden Fields optimization (listing \ref{list:Forbidden}) works by generating a list of forbidden fields. Those fields are creating deadlocks when the player pushes a box onto them. This approach has the benefit of having a lot of that one can exclude from all solution pathways. Those fields are deduced in the \#static part of the program, i.e. they are only grounded once and do not change over the timesteps in a level instance.
The disadvantage of this approach is the rather long time it takes to find all forbidden fields. Furthermore the approach works best in levels that do not have any restrictions on how they are constructed. As I am looking into levels having a very specific structure, see \ref{chapter 1 or 2 introduction}, this implementation is not the best approach. In rectangular levels there is an easier way to express a very similar deadlock detection, see next paragraph.

@@

My third approach to deadlock detection (see listing \ref{list:Clutter}) recognizes a "clutter" of boxes (see Fig. \ref{fig:clutter}). This means a 2x2 square where each field is either a box or a wall, which means that all those boxes would have to stay in their places. In line 8 of the optimization we can see that the box D does not have a target underneath which qualifies the situation to be a deadlock. This type of deadlock happens a lot in crammed levels with a high box to field ratio. Those very common deadlocks are clearly most important to find and remove to reduce the search tree efficiently.

@@

The optimization in listing \ref{list:MOOB} is very simple but can remove a lot of moves from grounding, which would be removed after ramifications when checking integrity constraints otherwise. This is more important in levels with a very long shortest plan. The optimization works by preventing all moves that would lead the player of the playable area, for instance a move to the left while the player is at the leftmost position of the level.

>>>>

@@

\section{Rules} \label{IntroStr}

@@

In my work I decided to research Sokoban planning with Answer Set Programming (ASP), a logic programming system designed to return answer sets to queries \cite{LifschitzASP}. I created different basic encodings and optimizations in a modular fashion. Those I will present to you in this chapter. Additionally I created a level portfolio, corresponding to different qualitative properties, e.g. level dimensions or number of boxes, and quantitative properties e.g. a seperation of the level into a zone where all the boxes are placed and a zone where all the targets are placed. The levels have two different types of structures. The first type is a rectangular level with the X and Y dimension differing by 1. The second type is created by connecting 2x2 fields at their edges, see \ref{fig:levelCrooked}.
%. But instead of Prolog ASP is capable of producing answer sets to a query

@@

For illustration consider the level shown in Figure \ref{fig:levelVis}. The corresponding fact representation can be seen in Listing \ref{lst:lvl1}. The board size is fixed by the constants $dimX$ and $dimY$ and the origin (1,1) is in the top left corner. The positions of player and boxes are stored in the at/3 predicate. Here the first argument represents the identifier. A zero stands for the player, while all boxes have numbers greater than zero. The second and third argument are the X and Y coordinates, signifying the initial position of the object. For instance init(at(0,1,2)) means the player is at the initial position (1,2). Because I am not exclusively looking at rectangular levels I decided to store all legal fields (meaning fields where a box or player can be placed) by using the predicate field/2. In rectangular levels the legal fields can be defined in one line. Here init(field(1..5,1..4)) (l.4) means the crossproduct of all the numbers in the pools (i.e. (1,1),..,(5,1),(1,2),..,(5,4)).

@@

\subsection{Split Encoding}
\lstinputlisting{../EncodingsClean/SokobanSplitEncoding.lp}
\begin{lstlisting}[caption={Split Encoding},label=list:SplitEnc]
\end{lstlisting}
The split encoding uses a technique common in ASP planning. When all coordinates are split into a vertical and a horizontal information the grounder does not have to ground all combinations of coordinates ($X * Y$) but only the sum of coordinates ($X + Y$). This has the advantage of less ground instances, but the disadvantage of a harder solving process. Just compare the ramifications in the plain encoding those in the split encoding. The plain encoding can deal with all ramifications in just three rules, while the split encoding needs ten rules. Obviously when splitting the coordinates into horizontal and vertical the rules deriving the new coordinates have to be twice as many. The other problem is that it is harder to find out whether two objects meet each other at a time. That is why the two frame axioms, signifying that boxes that are not touched do not move, have aggregates (ll.17-20). Those are used for simplifying complex rules; the first rule without the aggregates can be seen in listing \ref{list:BadFrame}.
\lstinputlisting{../EncodingsClean/BadSplitFrameAxiom.lp}
\begin{lstlisting}[caption={Frame Axiom X},label=list:BadFrame]
\end{lstlisting}
The last few changes are a second line for the initial readout of the level in lines 4 and 5, a few updates to the integrity constraints in lines 24 and 25 and a slightly different goal definition using the split coordinates.
%Here is another encoding, following a common ASP planning technique that aims to reduce ground instances. By splitting the horizontal and vertical coordinates, the program does not have to ground all combinations ($dimX * dimY$) of coordinates, but just each coordinate by itself (totaling $dimX + dimY$).% only grounds $(dimX + dimY) * horizon * N$ different positions instead of $dimX * dimY * horizon * N$.

@@

Whenever an AI tries to solve a planning problem with a naive encoding, it does not have the meta reasoning a human would have \cite{BoteaHeuristicsVsPlanning}, \cite{SokoRelevanceCuts}. This means it cannot see whether a move is worthless before it calculates its effect. So it has to play through all possible sequences of moves, even those where the first few moves already destroy any chance to satisfy the level. Those gamestates from which no solution can be reached are called deadlocks. A search tree can be greatly reduced if one were to remove all deadlocked branches. The difficulty is that some deadlocks are very hard to notice and sometimes it takes more time eliminating the deadlock than to explore the search tree without eliminating. This is why my deadlock detection approach removes only deadlocks, that are easy to check for. There are three different approaches I tried to remove deadlocks. The Forbidden Fields optimization (listing \ref{list:Forbidden}) works by generating a list of forbidden fields. Those fields are creating deadlocks when the player pushes a box onto them. This approach has the benefit of having a lot of fields that one can exclude from all solution pathways. Those fields are deduced in the \#static part of the program, i.e. they are only grounded once and do not change over the timesteps in a level instance.
The disadvantage of this approach is the rather long time it takes to find all forbidden fields. Furthermore the approach is suited for levels that do not have any restrictions on how they are constructed. As I am looking into levels having a very specific structure (see \ref{IntroStr}), this implementation is not the best approach. In rectangular levels there is an easier way to express a very similar deadlock detection, see next paragraph.

@@

My third approach to deadlock detection (see listing \ref{list:Clutter}) recognizes a "clutter" of boxes (see Fig. \ref{fig:clutter}). This is a 2x2 square where each field is either a box or a wall, which would imply that all those boxes would have to stay in their places. In line 8 of the optimization we can see that the box D does not have a target underneath, which qualifies the situation to be a deadlock. This type of deadlock happens a lot in crammed levels with a high box to field ratio. Those very common deadlocks are very important to find and remove to reduce the search tree in those levels efficiently.

@@

The optimization in listing \ref{list:MOOB} is very simple but can remove a lot of moves from grounding, which would be removed after ramifications when checking integrity constraints otherwise. This is more important in levels with a high horizon constant, i.e. a long shortest plan. The optimization works by preventing all moves that would lead the player of the playable area, for instance a move to the left, while the player is at the leftmost position of the level.
